Tree growth, mortality, and climatic stress in west-coast states, USA

### 2025-11-14 Manuscript.Rmd
Fourth version of the manuscript; edited the app slightly to make species/type of data shown easier to switch between and added mountain range names.

Numerous tree species in western North America exhibit increased mortality or decline.  van Mantgem et al. (2009) modeled mortality rates for the genera *Abies* and *Pinus* in the western USA as increasing from less than 0.05% annually in 1960 to almost 0.2% by 2010. Johnston et al. (2025) documented a 29% 10-year mortality rate of 300+ yr old trees in Eastern Oregon.  Perret et al. (2023) estimated a decadal range-wide mortality rate of 25% for Subalpine fir (*Picea engelmannii*).  High-severity fires in California in 2020 and 2021 killed 13%-19% of extant giant sequoia (*Sequoiadendron gianteum*) with >1.07-m bole diameters (Stephenson and Brigham 2021, Shive et al. 2022); recruitment in these burns was depressed (Soderberg et al. 2022).   Five of eight common species in western North America exhibit declines as measured by relative stand density (Stanke et al. 2021).  Shriver et al. (2022) found that four of five juniper (*Juniperus* spp.) and piñon pine (*Pinus* spp.) species were declining in portions of their range, comparing mortality to recruitment rates.

Climate change as manifest through increases or intensification in drought and aridity can affect tree mortality directly through physiological damage and indirectly by increasing wildfire frequency and severity and by promoting insect and disease outbreaks.  Hammond et al. (2022) found globally that the intensity of drought and heat events are directly associated with widespread forest die-off events.  Wildfire frequency and area burned in the western United States has increased, in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt (Abatzoglou and Williams 2016, Westerlin 2016, Juang et al. 2022).  There has also been a regional increase in the area of bark beetle (subfamily *Scolytinae* spp.) outbreaks (Meddens et al. 2012, Hicke et al. 2016).  Many of these indirect climate-influenced mortality causes may be attributable to the physical condition of trees following abnormal or prolonged exposure to water stress.

@@

One regionally useful metric of water stress is Climatic Water Deficit (CWD; Stephenson 1998).  Stephenson (1990) defines CWD as evaporative demand not met by available water, or the difference between plant water demand (actual evapotranspiration, AET), and plant water supply (potential evapotranspiration); these variables are driven by temperature and precipitation, respectively.  The Anderegg  et al. (2015) tree mortality study relied on CWD to capture drought occurrence.  Das et al. (2013) found low-elevation California tree mortality best modeled using CWD.  Restaino et al. (2016) found decreased Douglass fir growth with an increase in CWD.

Data from the U.S. National Forest Inventory, collected by the US Forest Service Forest Inventory and Analysis Program (FIA), is well suited for evaluating regional species mortality and growth.   Since 1999 the FIA program has gathered a variety of forestry data at plots nationwide using a spatially balanced probability sample (McRoberts 2005, Reams et al. 2005, Woudenberg et al. 2010).  The sample allows for a relatively unbaised design-based inference of population characteristics across a large geographical region.  Data from plot revisits and individual tree data permit direct estimation of change features such as mortality and growth estimation.

Climatic data such as mean annual temperature and CWD can be combined with FIA data to obtain climate-related design-based estimates. Monleon and Linz (2015) evaluated the evidence of tree species range shifts by subtracting the design-based mean estimates of annual FIA plot temperatures for seedlings from those of trees. FIA plot revisit data were mostly incomplete at the time of their analysis.  Groom and Monleon (2023) used a mostly-completed revisit data set to compare plot climate variable values (vapor pressure deficit [VPD], temperature, and precipitation) to estimate shifts in plot occupancy and density.  Both studies used a single climate variable value associated with each plot for estimation; e.g., Groom and Monleon (2023) examined shifts in species density by estimating the difference in VPD means between plots with more trees of a given species and plots with fewer trees.  They documented, however, that the initial climate variable values could (and often did) change between plot visits, with species often exhibiting positive or negative correlations between the initial visit and between-visits change value of the climate variable.  They identified that their design-based analysis could potentially improve by including both initial visit and change values for the climate variable.

The purpose of this analysis is to obtain design-based estimates of tree species’ growth and mortality across the western U.S. states of California, Oregon, and Washington as they relate to plot-level initial-visit water stress (initial CWD) and the change in water stress between plot visits (delta CWD).  We approached this bivariate analysis by dividing species’ plot data into domains based on initial and delta CWD values, finding growth and mortality estimates for each domain, and estimating domain contrasts to establish differences between domains.  Our secondary goal includes providing decadal growth and mortality estimates for our analysis species overall and by state to establish species status and to inform our CWD analysis findings.

@@

FIA plots consist of four points where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied in a 7.32 m radius circular subplot (total area 672.5 m$^2$) with additional trees sampled in an anulus around each subplot (macroplot), with a total radius 18 m from the center point and a total area of 1012 m$^2$.  We did not include data from microplots, where seedlings and saplings are measures, as the analysis was not intended to estimate species recruitment but rather to follow the trajectory of established trees.  Bechtold and Scott (2005) provide plot design details and measurement protocols.

The FIA program selected plots with a spatially-balanced probability sampling design (McRoberts 2005, Reams et al., 2005) that used a target plot density of one ground plot every 24 km$^2$. Some regions and ownerships intensify the plot density; we excluded these additional plots to simplify stratification estimation, relying on FIA-defined strata and using the density of FIA Phase 1 sampling points within and across strata to determine stratum weights.

@@

We estimated, for each species examined, six domains based on CWD values (Figure 2C).   The initial CWD plot values are divided into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High).  Delta CWD plot values for all species were divided into these above a 5mm CWD change threshold (Above) and those below the threshold (Below). The 5mm threshold was selected because it is approximately the mean delta CWD value across species and because species' delta CWD value distributions did not differ greatly (Table S2).  The category Below includes negative delta CWD values (i.e., mean CWD was larger prior to the initial visit than between visits).  The domain designations AL, AM, and AH correspond to Above Low, Above Medium, and Above High. The Below domains are similarly labeled (BL, BM, BH). We also report species mortality and growth estimates for all plots across and within Oregon, Washington, and California irrespective of CWD values.  We limited the number of species examined to those with more than 400 occupied plots across the three states.  We calculated domain estimates for all CWD-based domains with 10 or more plots.

For each species we examined contrasts among growth (Figure `r fig_nums("fig_diffgrow", display = "num")`) and mortality (Figure `r fig_nums("fig_diffmort", display = "num")`) CWD domain estimates.  The figure column "Above vs. Below Threshold" displays the mean point estimate and 95% CI for the contrast between corresponding Above and Below domains (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns BL, BM, and BH from AL, AM, AH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A). Negative estimates for the "Above vs. Below Threshold" in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`  indicate that the 10-year growth or mortality rate was greater for a species in plots below the change estimates.  `r t.ref("tbl_summ_grow_mort")` summarizes the number of domain differences in each of the columns in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` where the 95% CI does not include zero (termed "significant" findings) as well as the number of contrasts per figure column.

@@

Our analysis approach results in many comparisons.  We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates (p-values) for "significant" contrasts. The MC permutation analysis had 1000 iterations; each iteration estimated each species' bootstrap confidence interval as described above. The plot domain assignments for each species were sampled without replacement.  Given this sampling approach, and since species distributions were not independent, the p-values are strictly interperetable at the species level (e.g., we cannot speak to the likelihood of observing a specific number of significant findings for a contrast type). The distance of MC permutation contrast CI bounds from zero were compared against test CI bounds. The MC permutation procedure recorded a null significant value if the absolute values of the permutation iteration CI bounds were at least as far from zero as observed bounds.

## Dashboard
We developed a data dashboard (https://tinyurl.com/FIAgrowmort) to allow readers to examine, for individual species, a species' domain estimates, the distribution of initial and delta CWD values, and the geographical distribution of plots and plot domains.  All analysis files, manuscript markdowns, dashboard code, and ancillary data sources are available at the GitHub repository https://github.com/jeremygroom/Tree_Growth_Death.

@@

Mean overall growth rates for species varied between `r min.grow.spp$mean` (*`r min.grow.spp$spp`*) and `r max.grow.spp$mean` (*`r max.grow.spp$spp`*, `r t.ref("tbl_growth")`) cm$^2$/decade.  Six species had greater growth (non-overlapping CIs) in California than in Oregon and/or Washington: *Abies concolor*, *Abies grandis*, *Pinus ponderosa*, *Tsuga heterophylla*, *Arbutus menziesii*, and *Notolithocarpus densiflorus*.  Three species had greater growth in Oregon (*Calocedrus decurrens*, *Picea engelmanii*, and *Pseudotsuga menziesii*) and two species had greater growth in Washington (*Abies amablis* and *Pinus monticola*).  Decadal mean mortality rates varied between `r min.mort.spp$mean` (*`r min.mort.spp$spp`*) and `r max.mort.spp$mean` (*`r max.mort.spp$spp`*, `r t.ref("tbl_mort")`). California had four species with greater mortality rates than Oregon and/or Washington: *Abies concolor*, *Pinus ponderosa*, *Acer macrophyllum*, and *Quercus kelloggii*.  Oregon had one species with a greater mortality rate (*Tsuga mertensiana*) and Washington had four (*Abies grandis*, *Larix occidentalis*, *Picea engelmannii*, and *Pseudotsuga menziesii*).  Seven species had overall mortality rates $\geq$ 0.20 (*Abies concolor*, *Abies lasiocarpa*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Alnus rubra*, and *Quercus chryselopis*) while at a state level seven species had mortality rates $\geq$ 0.25 (California: *Acer macrophyllum* and *Alnus rubra*; Oregon: *Abies lasiocarpa*; Washington: *Abies grandis*, *Picea engelmannii*, *Pinus contorta*, and *Pinus monticola*).

@@

The MC permutation analysis provided empirical p-values for observing contrast confidence intervals with bounds at least as far from zero as those observed (Supplemental figures S5 and S6), with values ranging from a maximum of $p$ = 0.072 to a minimum $p$ < 0.0001 (41 instances).  Except where explicitly stated, we only discuss individual species' domain findings if the species has one or more relevant MC permutation p-values < 0.001 (indicated by three or four asterisks in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`)

@@

Conspecifics *Abies lasiocarpa* and *Picea engelmannii* are frequently codominant and exhibit high mortality rates in our study. Their respective mortality rates in our analysis appear to be elevated; Andrus et al. (2018) found that *Picea engelmannii* exhibited higher survival and longevity compared to *Abies lasiocarpa* in undisturbed plots in Colorado, with firs exhibiting a mortality rates of up to 1% annually while *Picea engelmannii* had mortality rates below 0.5% annually.  Mortality estimates for *Abies lasiocarpa* from Perret et al. (2023) approximated our state and overall estimates at 25% +/-2.7, with mortality concentrated in the eastern Oregon Cascades as well as in Idaho and Colorado. They found that 83% of the mortality was due to fire or biogenic disturbance.  The pressures on them appear to differ by CWD values and location.  We found that *Picea engelmannii* had lower mortality in the Above High domain, corresponding to stands in the Blue Mountains of Oregon and Washington, and higher mortality in the Above Low and Above High domains, largely coinciding with northern and northeastern Washington, where spruce beetle were elevated (WADNR 2023).  Its mortality rate in Washington was almost 5 times higher than in Oregon.  *Abies lasiocarpa* did not have substantial mortality contrast findings (some evidence of BH - BL difference, but the MC permutation p-value was greater than 0.001) and its mortality was high but similar across all three states.

Our findings for *Pinus contorta* in Washington likely reflected insect pressures on the species.  Our Washington mortality estimate was twice that of Oregon’s and five times greater than Calfornia’s, while the growth confidence intervals for Washington were lower than California’s and did not overlap. Mortality was highest in the Low CWD domains and growth lowest in the Above Low domain and also high in the Below Low and Below Medium domains.  These domains aligns well with plots for Washington, in regions that have experienced water stress (Raymond et al. 2023).  Pine bark beetles (*Dendroctonus ponderosae*, *Dendroctonus brevicomis*, and *Ips* spp.) have affected *Pinus contorta* for the past decade in this region (WADNR 2023).  This area also corresponds, at the level of ecoregion subsections, with the portions of *Pinus contorta*’s range in the lowest forest stability index category for Stanke et al. (2021) while California’s Sierra Nevadas and the Oregon and Washington Cascade Range correspond to positive stability index values.  Stanke et al. also found, from FIA plot data, that insects (followed by fire) were related to the percent change in relative tree density.

*Abies grandis* exhibited higher mortality in Washington than Oregon or California. We found no evidence of a difference in domain mortality geographically, although the greatest growth domain was Below Low, which corresponded to the more mesic plots west of the Cascade Crest of Washington and Oregon.  The mortality result may be due to an intersection in a fire suppression history promoting a densification of shade-tolerant but drought-intolerant firs in the arid climates of eastern Oregon and Washington, drought, and fir engraver beetle mortality (*Scolytus ventralis*; Washington DNR 2023, Merschel et al. 2021).  Johnson et al. (2025) report mortality rates of *Abies grandis* in protected areas in the Malheur National Forest of Oregon that exceed ours, primarily attributing fire as the cause.

@@

We did not detect sizable changes in CWD between visits, potentially signalling that 10 years is too soon to observe CWD-related species responses (perhaps moreso for mortality than growth).  The CWD changes we observed may have represented greater proportional change for some species than others, with species such as *Abies amablis*, *Alnus rubra*, and *Juniperus occidentalis* exhibiting narrow and low water stress initial CWD ranges. CWD effects can manifest in direct and indirect ways, potentially differing with species by geographic location and the biotic and abiotic impacts to those locations.  It also cannot easily convey differences in impacts to trees within the same stands.

@@

Our analysis may overall capture significant changes in mortality and growth for the tree species examined, with some trees exhibiting increases in mortality and decreases in growth either in portions of their ranges or across their entire range.  For some species in semi-arid environments such as *Pinus pondersosa*, *Pseudotsuga menziesii*, *Pinus contorta*, *Picea engelmannii*, and *Abies lasiocarpa*, we may be observing ecosystem shifts with a change in water stress combined with disturbances such as wildfire and insect mortality (Halofsky et al. 2020, Parks et al. 2019, Raymond et al. 2023).  Our hope is that our bivariate domain-based estimation approach is useful for others, while our results for individual species serve bolster or direct others' inquiries into the status of the species we examined.  For managers, the results may be used to bolster action for resiliency.   With the shift in ecosystems, landowners can assist with species migration when planting or replanting (Raymond et al. 2023), and forest thinning can help reduce water stress and the impacts of fire and insects (Bernal et al. 2023).

@@

Our analysis serves as an example of bivariate design-based estimation, where we use estimation domains based on initial CWD and delta CWD values of plots.  For the 23 species we examined, we found no general patterns in domains with higher or lower tree growth or mortality; the results appeared species specific.  We believe our analysis approach can be refined for individual species analysis, with the potential for integrating a third variable that could introduce information besides a condition variable and change variable.  Our analysis and the analyses of others provide valuable and reliable insights into the condition of several tree species that exhibit abnormally high mortality (and low growth).  Our hope si that these findings will help drive management and further monitoring of these species in the rapidly changing landscape of western North America.

>>>>

Numerous tree species in western North America exhibit increased mortality rates or decline.  van Mantgem et al. (2009) modeled mortality rates for the genera *Abies* and *Pinus* in the western USA as increasing from less than 0.05% annually in 1960 to almost 0.2% by 2010. Johnston et al. (2025) documented a 29% 10-year mortality rate of 300+ yr old trees in Eastern Oregon.  Perret et al. (2023) estimated a decadal range-wide mortality rate of 25% for Subalpine fir (*Picea engelmannii*).  High-severity fires in California in 2020 and 2021 killed 13%-19% of extant giant sequoia (*Sequoiadendron gianteum*) with >1.07-m bole diameters (Stephenson and Brigham 2021, Shive et al. 2022); recruitment in these burns was depressed (Soderberg et al. 2022).   Five of eight common tree species in western North America exhibit declines as measured by relative stand density (Stanke et al. 2021).  Shriver et al. (2022) found that four of five juniper (*Juniperus* spp.) and piñon pine (*Pinus* spp.) species were declining in portions of their range, comparing mortality to recruitment rates.

Climate change as manifest through increases or intensification in drought and aridity can affect tree mortality directly through physiological damage as well as indirectly by increasing wildfire frequency and severity and by promoting insect and disease outbreaks.  Hammond et al. (2022) found globally that the intensity of drought and heat events are directly associated with widespread forest die-off events.  Wildfire frequency and area burned in the western United States has increased in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt (Abatzoglou and Williams 2016, Westerlin 2016, Juang et al. 2022).  There has also been a regional increase in the area of bark beetle (subfamily *Scolytinae* spp.) outbreaks (Meddens et al. 2012, Hicke et al. 2016).  Many of these indirect climate-influenced mortality causes may be attributable to the physical condition of trees following abnormal or prolonged exposure to water stress.

@@

One regionally useful metric of water stress is Climatic Water Deficit (CWD; Stephenson 1998).  Stephenson (1990) defines CWD as evaporative demand not met by available water, or the difference between plant water demand (actual evapotranspiration, AET), and plant water supply (potential evapotranspiration); these variables are driven by temperature and precipitation, respectively.  The Anderegg  et al. (2015) tree mortality study relied on CWD to capture drought occurrence.  Das et al. (2013) determined that low-elevation California tree mortality was best modeled using CWD.  Restaino et al. (2016) found decreased Douglass fir growth with an increase in CWD.

Data from the U.S. National Forest Inventory, collected by the US Forest Service Forest Inventory and Analysis Program (FIA), is well suited for evaluating regional species mortality and growth.   Since 1999 the FIA program has gathered a variety of forestry data at plots nationwide using a spatially balanced probability sample (McRoberts 2005, Reams et al. 2005, Woudenberg et al. 2010).  The sample allows for relatively unbiased design-based inference of population characteristics across a large geographical region.  Data from plot revisits and individual tree data permit direct estimation of change features such as mortality and growth estimation.

Climatic data such as mean annual temperature and CWD can be combined with FIA data to obtain climate-related design-based estimates. Monleon and Linz (2015) evaluated the evidence of tree species range shifts by subtracting the design-based mean estimates of FIA plot temperatures for seedlings from those of trees. FIA plot revisit data were mostly incomplete at the time of their analysis.  Groom and Monleon (2023) used a mostly-completed revisit data set to compare plot climate variable values (vapor pressure deficit [VPD], temperature, and precipitation) to estimate shifts in plot occupancy and density.  Both studies used a single climate variable value associated with each plot for estimation; e.g., Groom and Monleon (2023) examined shifts in species density by estimating the difference in VPD means between plots that gained trees of a given species and plots that lost trees.  They documented, however, that the initial climate variable values could (and often did) change between plot visits, with species' plots often exhibiting positive or negative correlations between the initial visit values and between-visits change value of the climate variable.  They identified that their design-based analysis would potentially improve if they could simultaneously include both initial visit and change values for the climate variable.

The purpose of this analysis is to demonstrate an approach for a bivariate design-based estimation procedure. We estimate tree species’ growth and mortality across the western U.S. states of California, Oregon, and Washington as they relate to plot-level initial-visit water stress (initial CWD) and the change in water stress between plot visits (delta CWD).  We approached this bivariate analysis by dividing species’ plot data into domains based on initial and delta CWD values, finding growth and mortality estimates for each domain, and estimating domain contrasts to establish differences between domains.  Secondarily we provide decadal growth and mortality estimates for our analysis species overall and by state to establish species status and to inform our CWD analysis findings.

@@

FIA relies on three related sampling procedures.  Phase 1 involves the classification of distributed points of remote sensing data into (at least) the categories of forest and non-forest.  Phase 1 plots are useful for estimation stratification.  The phase 2 permanent ground plots are a subset of Phase 1 points.  They consist of four points where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied in a 7.32 m radius circular subplot (total area 672.5 m$^2$) with additional trees sampled in an annulus around each subplot (macroplot), with a total radius 18 m from the center point and a total area of 1012 m$^2$.  We did not include data from microplots, where seedlings and saplings are measured, as the analysis was not intended to estimate species recruitment but rather to follow the trajectory of established trees. Phase 3 plots are an intensification of Phase 2 plots; we do not rely on their data here. Reams et al. (2005) and Bechtold and Scott (2005) provide plot design details and measurement protocols.

The FIA program selected plots with a spatially-balanced probability sampling design (McRoberts 2005, Reams et al., 2005) that used a target plot density of one ground plot every 24 km$^2$. Some regions and ownerships intensify their plot density; we excluded these additional plots to simplify stratification estimation, relying on FIA-defined strata and using the density of FIA Phase 1 sampling points within and across strata to determine stratum weights.

@@

We estimated, for each species examined, six domains based on CWD values (Figure 2C).   The initial CWD plot values are divided into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High).  Delta CWD plot values for all species were divided into those above a 5mm CWD change threshold (Above) and those below the threshold (Below). The 5mm threshold was selected because species' delta CWD value distributions did not differ greatly and because 5mm is approximately the mean delta CWD value across species (Table S2).  The Below domains may includes negative delta CWD values (i.e., mean CWD was larger prior to the initial visit than between visits).  The domain designations AL, AM, and AH correspond to Above Low, Above Medium, and Above High. The Below domains are similarly labeled (BL, BM, BH). We also report species mortality and growth estimates for all plots across and within Oregon, Washington, and California irrespective of CWD values.  We limited the number of species examined to those with more than 400 occupied plots across the three states.  We calculated estimates for all CWD-based domains with 10 or more plots.

For each species we examined contrasts among growth (Figure `r fig_nums("fig_diffgrow", display = "num")`) and mortality (Figure `r fig_nums("fig_diffmort", display = "num")`) CWD domain estimates.  The figure column "Above vs. Below Threshold" displays the mean point estimate and 95% CI for contrast between corresponding Above and Below domains (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns BL, BM, and BH from AL, AM, AH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A). Negative estimates for the "Above vs. Below Threshold" in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`  indicate that the 10-year growth or mortality rate was greater for a species in plots below the change estimates.  `r t.ref("tbl_summ_grow_mort")` summarizes the number of domain differences in each of the columns in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` where the 95% CI does not include zero (termed "significant" findings) as well as the number of contrasts per figure column.

@@

Our analysis approach results in many comparisons.  We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates (p-values) for "significant" contrasts. For each of 1000 MC permutation analysis iterations the code estimated each all species' bootstrap confidence intervals as described above. The MC permutation analysis sampled each species' plot domain assignments without replacement.  Given this sampling approach, and since species distributions were not independent, the p-values are strictly interpretable at the species level (e.g., we cannot speak to the likelihood of observing a specific number of significant findings for a contrast type). The distance of MC permutation contrast CI bounds from zero were compared against test CI bounds. The MC permutation procedure recorded a null significant value if the absolute values of the permutation iteration CI bounds were at least as far from zero as observed bounds.

## Dashboard
We developed a data dashboard (https://tinyurl.com/FIAgrowmort) to enable ourselves and readers to examine individual species domain estimates, the distribution of initial and delta CWD values, and the geographical distribution of plots and plot domains.  All analysis files, manuscript markdowns, dashboard code, and ancillary data sources are available at the GitHub repository https://github.com/jeremygroom/Tree_Growth_Death.

@@

Mean overall growth rates for species varied between `r min.grow.spp$mean` (*`r min.grow.spp$spp`*) and `r max.grow.spp$mean` (*`r max.grow.spp$spp`*, `r t.ref("tbl_growth")`) cm$^2$/decade.  Six species had greater growth (non-overlapping CIs) in California than in Oregon and/or Washington: *Abies concolor*, *Abies grandis*, *Pinus ponderosa*, *Tsuga heterophylla*, *Arbutus menziesii*, and *Notolithocarpus densiflorus*.  Three species had greater growth in Oregon (*Calocedrus decurrens*, *Picea engelmanii*, and *Pseudotsuga menziesii*) and two species had greater growth in Washington (*Abies amablis* and *Pinus monticola*).  Decadal mean mortality rates varied between `r min.mort.spp$mean` (*`r min.mort.spp$spp`*) and `r max.mort.spp$mean` (*`r max.mort.spp$spp`*, `r t.ref("tbl_mort")`). Four species in California exhibited greater mortality rates than in Oregon and/or Washington: *Abies concolor*, *Pinus ponderosa*, *Acer macrophyllum*, and *Quercus kelloggii*.  One species had a greater mortality rate in Oregon (*Tsuga mertensiana*) and Washington had five (*Abies grandis*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, and *Pseudotsuga menziesii*).  Seven species had overall mortality rates $\geq$ 0.20 (*Abies concolor*, *Abies lasiocarpa*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Alnus rubra*, and *Quercus chryselopis*) while at a state level seven species had mortality rates $\geq$ 0.25 (California: *Acer macrophyllum* and *Alnus rubra*; Oregon: *Abies lasiocarpa*; Washington: *Abies grandis*, *Picea engelmannii*, *Pinus contorta*, and *Pinus monticola*).

@@

The dashboard revealed domains of interest ("significant") for certain species that aligned with geographical regions.  For the Above and Below Low domains *Abies lasiocarpa* had less mortality in northeastern Washington and *Tsuga mertensiana* exhibited greater growth in the same region.  *Pinus contorta* exhibited less growth and greater mortality in northern Washington (Above Low).  In the southern Sierra Nevada *Pinus lambertiana* had greater growth (Above Medium and High) and greater mortality (Above High), *Calocedrus decurrens* had greater mortality (Above High), and *Pinus monticola* had less mortality (Above High). *Acer macrophyllum* exhibited less growth in the Sierra Nevada and southern Klamath Mountains (Above High) and greater growth in northwestern Washington in the Coast and Cascade ranges (Above and Below Low).

The MC permutation analysis provided empirical p-values for observing contrast confidence intervals with bounds at least as far from zero as those observed (Supplemental figures S5 and S6), with values ranging from a maximum of $p$ = 0.072 to a minimum $p$ < 0.0001 (41 instances).  Except where explicitly stated, we only discuss individual species' domain findings if the species has one or more relevant MC permutation p-values < 0.001 (indicated by three or four asterisks in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`).

@@

Conspecifics *Abies lasiocarpa* and *Picea engelmannii* are frequently codominant and exhibit high mortality rates. Andrus et al. (2018) found that *Picea engelmannii* exhibited higher survival and longevity compared to *Abies lasiocarpa* in undisturbed plots in Colorado, with firs exhibiting a mortality rates of up to 1% annually while *Picea engelmannii* had mortality rates below 0.5% annually.  Mortality estimates for *Abies lasiocarpa* from Perret et al. (2023) approximated our state and overall estimates at 25% +/-2.7; they found the highest level of mortality in the eastern Oregon Cascades as well as in Idaho and Colorado. They attributed 83% of the mortality to fire or biogenic disturbance.  We found that the pressures on the two species appear to differ by CWD values and location.  We found that *Picea engelmannii* had lower mortality in the Above High domain, corresponding to stands in the Blue Mountains of Oregon and Washington, and higher mortality in the Above Low and Above High domains, largely coinciding with northern and northeastern Washington, where spruce beetle were elevated (WADNR 2023).  Its mortality rate in Washington was almost 5 times higher than in Oregon.  *Abies lasiocarpa* did not have substantial mortality contrast findings (some evidence of BH - BL difference, but the MC permutation p-value was greater than 0.001) and its mortality was high but similar across all three states.

Our findings for *Pinus contorta* in Washington likely reflected insect pressures on the species.  Our Washington mortality estimate was twice that of Oregon’s and five times greater than Calfornia’s, while the growth confidence intervals for Washington were lower than California’s and did not overlap. The domains with the highest mortality and lowest growth align well with Washington regions that have experienced water stress (Raymond et al. 2023).  Pine bark beetles (*Dendroctonus ponderosae*, *Dendroctonus brevicomis*, and *Ips* spp.) have affected *Pinus contorta* for the past decade in this region (WADNR 2023).  This area also corresponds, at the level of ecoregion subsections, with the portions of *Pinus contorta*’s range in the lowest forest stability index category for Stanke et al. (2021) while California’s Sierra Nevada and the Oregon and Washington Cascade Range correspond to positive stability index values.  Stanke et al. also found, from FIA plot data, that insects (followed by fire) were related to the percent change in relative tree density.

*Abies grandis* exhibited higher mortality in Washington than Oregon or California. We found no evidence of a difference in domain mortality geographically, although the greatest growth domain was Below Low, which corresponded to the more mesic plots west of the Cascade Crest of Washington and Oregon.  The mortality result may be due to an intersection in a fire suppression history promoting a densification of shade-tolerant but drought-intolerant firs in the arid climates of eastern Oregon and Washington, drought, and fir engraver beetle mortality (*Scolytus ventralis*; Washington DNR 2023, Merschel et al. 2021).  Johnson et al. (2025) report mortality rates of *Abies grandis* in protected areas in the Malheur National Forest of Oregon that exceed ours, primarily attributing the cause to fire.

@@

We did not detect sizable changes in CWD between visits, potentially signalling that 10 years is too soon to observe CWD-related species responses (perhaps more true for mortality than growth).  The CWD changes we observed may have represented greater proportional change for some species than others, with species such as *Abies amablis*, *Alnus rubra*, and *Juniperus occidentalis* exhibiting narrow and low water stress initial CWD ranges. CWD effects can manifest in direct and indirect ways, potentially differing with species by geographic location and the biotic and abiotic impacts to those locations.  It also cannot easily convey differences in impacts to trees within the same stands.

@@

Our analysis may overall capture significant changes in mortality and growth for the tree species examined, with some trees exhibiting increases in mortality and decreases in growth either in portions of their ranges or across their entire range.  For some species in semi-arid environments such as *Pinus ponderosa*, *Pseudotsuga menziesii*, *Pinus contorta*, *Picea engelmannii*, and *Abies lasiocarpa*, we may be observing ecosystem shifts with a change in water stress combined with disturbances such as wildfire and insect mortality (Halofsky et al. 2020, Parks et al. 2019, Raymond et al. 2023).  Our hope is that our bivariate domain-based estimation approach is useful for others, while our results for individual species serve bolster or direct others' inquiries into the status of the species we examined.  For managers, the results may be used to inform action for resiliency.   With the shift in ecosystems, landowners can assist with species migration when planting or replanting (Raymond et al. 2023), and forest thinning can help reduce water stress and the impacts of fire and insects (Bernal et al. 2023).

@@

Our analysis serves as an example of bivariate design-based estimation, where we use estimation domains based on initial CWD and delta CWD values of plots.  For the 23 species we examined, we found no general patterns in domains with higher or lower tree growth or mortality; the results appeared species specific.  We believe our analysis approach can be refined for individual species analysis, with the potential for integrating a third variable that could introduce information besides a condition variable and change variable.  Our analysis and the analyses of others provide valuable and reliable insights into the condition of several tree species that exhibit abnormally high mortality (and low growth).  Our hope is that these findings will help drive management and further monitoring of these species in the rapidly changing landscape of western North America.
